An integrated open population distance sampling approach for modelling age-structured populations

### 2024-01-07 Manuscript_Nilsen_Nater.qmd
Submit version of manuscript + bib-file

---
title: "An integrated open population distance sampling approach for modelling age-structured populations"
format: html
author: Erlend B. Nilsen & Chloé R. Nater
editor: visual
bibliography: rypeIDSM_MS1.bib
---

## Abstract

*Estimation of abundance and demographic rates for populations of wild species is a challenging but fundamental issue for both management and research into ecology and evolution. One set of approaches that has been used extensively to estimate abundance of wildlife populations is Distance Sampling (DS) methods for line or point transect survey data. Historically, DS models were only available as closed* <!-- I think you mean "closed" here? --> *population models, and did not allow for the estimation of changes in abundance through time. The advent of open population formulations based on the DS framework greatly extended the scope of the models, but much untapped potential remains in models that estimate temporal dynamics not only in abundance but also in the underlying demographic rates. Here, we present an integrated distance sampling approach that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population dynamics and the demographic rates (recruitment rate and survival probability) that drive temporal changes . The core of the resulting model is, in essence, equivalent to both an integrated population model and a matrix population model with two age classes: juveniles and adults. The integrated framework allows making full use of the available data by effectively combining line transect and telemetry data, and can easily be adapted to include additional and/or different data types. Moreover, as demographic rates often respond to environmental variation, our approach also supports direct estimation of the effects of such environmental covariates on demographic rates. Through a comprehensive simulation study we show that the model is able to adequately recover true population and vital rate dynamics. Subsequent application to data from a study of willow ptarmigan (Lagopus lagopus) in Norway showcases the frameworks ability to recover both fluctuations and trends in population dynamics and highlights its potential applicability to a wide range of species surveyed using distance sampling approaches.*

@@

<!-- Just a note: we seems to also still be missing quite a few references in the introduction -->

Estimating abundance and demographic rates for wildlife populations is an integral part of basic and applied ecology [@Skalski2005, @williams2002]. Over the last few decades, tremendous progress has been made towards this end. This progress is partly driven the development and application of new field data collection methods and approaches, such as citizen science data (ref), camera trap data (ref) and the collection of environmental DNA data (ref). In addition, developments of novel statistical methods alongside decreases in computational costs now allow researchers to estimate abundance and demographic rates in situations where it was not feasible before (ref). Combined, these advances put us in a much better position for estimating quantities needed for population management [@williams2002] and indices relevant for large scale policy applications, e.g. Essential Biodiversity Variables [@Kissling2018].

Until recently, joint estimation of population dynamics and demography has relied mostly on data from marked individuals and associated open-population capture-mark-recapture models [@schaub2021]. While such methods can provide valuable information for both ecological research and management, collecting the necessary data is typically costly and logistically challenging to implement over large areas. Monitoring programmes focusing on abundance trends over larger areas, on the other hand, are typically based on data from unmarked animals (ref). A common approach for such surveys is to structure data collection around the distance sampling (DS) method. DS has been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015]. One reason for the method's popularity is that it does not require repeated visits to the same sites for estimating detection probability. This makes DS particularly useful for implementation in participatory monitoring programs, allowing stakeholders to take part in the data collection process.

The most common implementations of DS models have long used closed-population formulations, and thus do not allow formulating a process model that projects abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates @buckland2015. In recent years, DS approaches have been extended in many ways, including applications that estimate changes in abundance over time in open populations via a hidden state model representing population dynamics [@moore2011; @sollmann2015]. This has greatly extended the potential of DS approaches for studying ecological dynamics across time and space. However, while these latter frameworks may allow to accurately quantify population changes, they typically provide little information on the drivers of these changes, i.e. the underlying vital rates. In fact, if the data does not contain information about the age (and/or sex-) structure of the surveyed population, there is no straightforward way to estimate demographic rates from such data. On the contrary, if age (and sex) of detected individuals can be determined, this information can be used to provide information on recruitment rates and survival probabilities. @Nilsen2018, for example, used a model based on harvest statistics and observations of population age structure to estimate population abundance and demographic rates without the need for any additional data from marked individuals.

@@

<!-- Note: missing ref below. Maybe good to "search" the whole text for "ref" to make sure we have replaced the placeholder with appropriate references everywhere -->

@@

The willow ptarmigan has a circumpolar distribution (Fuglei et al., 2020), and lives year-round in heterogeneous alpine and artic ecosystems. In Norway, there has been a long-term decline in the willow ptarmigan abundance across more than a century (ref), but in the last few decades abundance trends has fluctuated both in time and space. In Scandinavia, willow ptarmigan is a valued game species (ref), and there have been several long-term research projects devoted to understanding how they respond to environmental variation and harvest management [@Israelsen2020; @Sandercock2011]. A key insight from across several study areas is the the annual recruitment rate (i.e. $R_t$ in our model, as outlined above) is highly variable, and is affected both by spring conditions [@Eriksen2023] and the abundance of small rodents, which constitute alternative prey for common predators (i.e. the Alternative Prey Hypothesis ref). Adult survival show less inter-annual fluctuations [@Israelsen2020], although spatial (and potentially temporal) variation due to e.g. harvest management is evident when comparing across studies [@Israelsen2020].

@@

Line transect survey data were collected in August each year, prior to the annual autumn harvest season, as part of the program "Hønsefuglportalen". Hønsefuglportalen is a national program for line transect surveys of tetraonid birds, and the effort is directed mainly towards willow ptarmigan habitats. In our case study, we used data from the western part of Lierne municipality. Line transects are surveyed by trained volunteers that use pointing dogs to locate the birds. When located, the geographical coordinate, perpendicular distance from the sampling line, the number of birds in the group, as well as the age (juvenile or adult) and sex of the birds are recorded. As the surveys are conducted in early August, juveniles can be distinguished from adults by their smaller body size. <!-- maybe also mention how males and females are distinguished? Also, is it just body size for the juveniles or also plumage? --> Nevertheless, mis-identification can occur, and in addition a proportion of the birds are registered as "unknown" age and/or sex. Since 2019, data has been collected through a mobile app tailormade for this project <!-- Add name + maybe link to download in app store? --> . Before 2019, field workers reported their data through a dedicated web portal. After data are collected and reported, they undergo several steps of quality control: first by local contacts and subsequently by personnel at the Norwegian Institute for Nature Research (NINA). The data are then standardized based on the Darwin-Core standard (ref), and made publicly available as a sampling-event data set published through GBIF [@Nilsen2023a]. For additional description of the data collection procedures, see [@Bowler2020]; [@Kvasnes2018][@Nilsen2023a].

The individual longitudinal study based on radio collared willow ptarmigan was conducted in 2015-2020. <!-- Above you stated that radio telemetry data collection is until 2021 and ongoing. Until what year was it? --> Each winter (in February-March), willow ptarmigan were located at night using snowmobiles and large hand nets with prolonged handles, as described in [@Israelsen2020]. To prevent birds from flying off before the field personnel were close enough to capture them, a high-powered head lamp was used to dazzle the birds. After capture, birds were placed in an opaque bag to reduce stress. They were then fitted with a uniquely numbered leg ring (\~ 2.4g) and a Holohil RI-2BM or Holohil RI-2DM radio transmitter (\~ 14.1g) and subsequently released. The radio transmitters had an expected battery lifetime of 24 months (RI-2BM) or 30 months (RI-2DM), and included a mortality circuit that was activated if a bird had been immobile for 12 hours. We monitored the birds throughout most of the year by triangulation from the ground at least once a month for 10 months of the year (February -- November) by qualified field personnel. If a mortality signal was emitted from a transmitter, we attempted to recover it as soon as possible to determine cause of death. A number of birds dispersed out of the main study areas and was thus out of signal range for field personnel on the ground. To avoid loss of data, we conducted aerial triangulation using a helicopter or airplane three times a year (May, September and November) in the years 2016-2020.

@@

For the model fits to simulated and real data we ran 3 and 4 MCMC chains with NIMBLE's standard samples for 500k and 100k iterations, respectively. 300k and 40k and thereof were discarded as burn-in prior to thinning with factors 5 and 20 , leaving us with 40k and 3k posterior samples per chain (total of 120k and 12k samples per run), respectively.

@@

Posterior distributions for parameters estimated in three model fits to each of 10 simulated data sets are shown in @fig-sim-detection and @fig-sim-demo. Overall, the IDSM was able to correctly estimate both detection parameters and demographic rate parameters from all 10 simulated datasets without any systematic bias. The replicate runs for each dataset resulted in very similar posterior distributions, demonstrating that the models converged towards the same posterior distributions irrespective of starting values. Estimates of population sizes / densities were also adequate, and are presented in the Supplementary Materials (see folder SimCheck_byDataSet).

@@

Given the available data, the IDSM was not able to estimate a clear effect of small rodent abundance on ptarmigan recruitment (slope-paramater for the z-standardized rodent occurrence data = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$Mean, 3)` ; 95% C.I. = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$lower, 3)` - `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$upper, 3)`, Supplementary Figure "Rep_betaR.R.png").

@@

We developed an integrated model that jointly analyses line transect distance sampling survey data and data from marked individuals to estimate population abundance, survival probabilities ,and recruitment rates over time. We first used simulated data to examine model's ability to recover the underlying parameters when they were known. We then fitted the model to data from an ongoing field study on willow ptarmigan in Norway to showcase its applicability to real wildlife monitoring data.

Open population formulations of the distance sampling model have previously been presented (refs), and applied to various ecological systems (refs). The model presented here extends those previous applications by formulating the underlying population process model as a stage-structured matrix model [@caswell2000] in which the matrix elements are represented by annual survival probabilities and recruitment rates. While this has been the common approach for a range of other statistical modelling frameworks, including the growing suite of models falling into the category of integrated population models [@schaub2022], the integration of mechanistic population models into distance sampling frameworks is rather new. The resulting modelling framework allow us to make maximum use of distance sampling data in combination with auxiliary information both from the distance sampling survey itself (i.e. information on age, sex, etc. of observed animals) and from other types of monitoring, and enables estimation not only of changes in population density but also of underlying vital rates over time.

@@

Real data, however, are likely to be subject to certain biases, and one type of bias that is likely to be common for age-structured distance sampling data arises from failure to (correctly) classify the age class of observed individuals. <!-- Maybe add one illustrative sentence here about how/why this happens using our ptarmigan monitoring setup as an example? --> Observations with incorrectly classified age have the potential to introduce bias in the IDSMs relative estimates of survival and recruitment. This is due to the way it usese the distance sampling data to estimate survival and recruitment rates. In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, and this creates a dependence between the demographic parameters. If the age ratio in the data are biased or contain frequent misclassifications, this is likely to affect the relative contribution of survival and recruitment to the growth rate. To get an idea of the potential effect of this on parameter estimation, we checked the sensitivity of the output of the model fit to real data with regard to the treatment of birds classified as "unknown sex and age" by the field personnel (see **Methods**). In the model version presented in the results section, we made the assumption that these birds were in fact juveniles. <!-- Restate here why this is likely / probably a good guess --> Comparing estimates to an alternative scenario in which we discarded all birds classified as "unknown sex and age" (Appendix) we found that -- as expected -- estimated population density was virtually unaffected by the treatment of "unknown sex and age" observations, while the annual demographic parameters shifted proportional to the amount of "unknown sex and age" observations in the given year (towards higher recruitment and lower survival). Thus, biases in the reported age ratios may affect estimated of demographic rates, but not so much population density. Since the proportion of "unknown sex and age" observations in our ptarmigan case study was rather low, potential biases in estimates resulting from age misclassification are expected to be small. Nonetheless, future developments of the IDSM modelling framework should focus on ways of accounting explicitly for misclassification of age class in the field.

<!-- I suggest we start this next paragraph with a couple of general sentences about the density estimates, even if it is just stating that the density estimates correspond closely to what we know/have estimated with other models previously. Maybe also add one sentence or so about the possible biology behind the recent increase. Then moving on to survival and recruitment -->

Compared to previous studies on ptarmigan (see e.g. [@Israelsen2020; @Sandercock2011]), we could have expected somewhat higher estimates of survival probability. One potential reason for the overall lower estimates obtained here is that our IDSM analysis assumed constant survival over the period 2007-2021 while survival, in reality, may have changed over time. If survival in more recent years, when telemetry data was collected [and the study of e.g. @@Israelsen2020 was carried out] was higher than in earlier years -- something that seems likely given population increase over recent years -- an average over the entire time period is expected to be lower. On a different note, we can also not exclude the possibility of a small degree of bias in survival estimates due to misclassification of age in the data (see above), especially seeing as the IDSM's recruitment rate estimates also appear somewhat high compared to other studies on ptarmigan (+refs). <!-- If we have any "biological" thoughts about why somewhat high recruitment rates may not be that ridiculous here, then that could be added here -->

In addition to estimating demographic rates from line transect data, the IDSM also allows including relevant environmental effects on the demographic rates themselves, and not just on population growth rate as a whole ($\lambda$).
In the ptarmigan case study we thus attempted to investigate the effect of small rodent abundance (approximated as the proportion of transect lines on which rodents were reported each year) on recruitment rate. <!-- Deleted this because we did NOT include the rodent effect in the simulated data tests --> Unfortunately, we were not able to detect a clear effect of rodent abundance due to large uncertainty associated with the estimate (Supplementary Figure "Rep_betaR.R.png"). This may seem somewhat surprising at first glance, given that such a pattern has been reported repeatedly in the literature (see e.g. @Bowler2020). We speculate that there are (at least) three potential explanations to this. The first is that our covariate data may not have been well suited for estimating effects on recruitment. The data on rodent abundance was heavily zero-inflated, and the annual variation in the index was rather small otherwise, making for a covariate with relatively little information content. While this may be partially a consequence of how these data are collected, it is also well known that the amplitude and regularity of the rodent cycles has been fading in recent decades (refs), and our study area might be no exception. Lack of peak rodent years in the time series to which we fitted the model may thus also have contributed to making effect estimation challenging. Second, it is possible that rodent effects were obscured by other, potentially stronger, covariate effects. Previous research has shown that ptarmigan recruitment is also sensitive to the weather in the late winter and spring (before and during the breeding season); as we did not fit any weather covariates to the model so there is a possibility that strong unaccounted for effects of spring conditions in certain years may have masked any remaining effects of small rodent abundance. Finally, the data set used in this analysis is relatively short (15 years), leaving us with somewhat limited statistical power to detect effects of temporal covariates . Taken together we therefore do not consider this study as a particularly strong test of the underlying effect of small rodent fluctuations on ptarmigan recruitment rates. It is worth noting that future applications could increase statistical power by including either more years of data or capitalizing on space-for-time substitution as the Norwegian ptarmigan monitoring programme spans many more locations beyond Lierne. @Bowler2020, for example, used data from the same sampling program but from more areas using a simpler open population DS model, and detected a very clear signal from small rodent abundance on ptarmigan population growth rate. An extension of our IDSM to include data from multiple areas therefore constitutes a promising approach for investigating to which extent similar results emerge by linking environmental covariates to the actual demographic rates and not only just to the resulting population growth rates.

@@

We are grateful to field workers that collected the line transect data through the Hønsefuglportalen program. Fjellstyrene i Lierne contributed both to the line transect program and the field work related to the marked willow ptarmigan. We also thank James A. Martin for constructive discussions during the later stages of this work. 
Both data collection and analytical work were financed by the Norwegian Environment Agency funded the projects (grant nrs **TO BE ADDED**).

@@

>>>>

---
title: "An integrated open population distance sampling approach for modelling age-structured populations"
format: docx
author: Erlend B. Nilsen & Chloé R. Nater
editor: visual
bibliography: rypeIDSM_MS1.bib
---

### Abstract

*Estimation of abundance and demographic rates for populations of wild species is a challenging but fundamental issue for both management and research into ecology and evolution. One set of approaches that has been used extensively to estimate abundance of wildlife populations is Distance Sampling (DS) methods for line or point transect survey data. Historically, DS models were only available as closed* *population models, and did not allow for the estimation of changes in abundance through time. The advent of open population formulations based on the DS framework greatly extended the scope of the models, but much untapped potential remains in models that estimate temporal dynamics not only in abundance but also in the underlying demographic rates. Here, we present an integrated distance sampling approach that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population density and the demographic rates (recruitment rate and survival probability) that drive temporal changes density. The core of the resulting model is, in essence, equivalent to both an integrated population model and a matrix population model with two age classes: juveniles and adults. The integrated framework allows making full use of the available data by effectively combining line transect and telemetry data, and can easily be adapted to include additional and/or different data types. Moreover, as demographic rates often respond to environmental variation, our approach also supports direct estimation of the effects of such environmental covariates on demographic rates. Through a comprehensive simulation study we show that the model is able to adequately recover true population and vital rate dynamics. Subsequent application to data from a study of willow ptarmigan (Lagopus lagopus) in Norway showcases the frameworks ability to recover both fluctuations and trends in population dynamics and highlights its potential applicability to a wide range of species surveyed using distance sampling approaches.*

@@

Estimating abundance and demographic rates for wildlife populations is an integral part of basic and applied ecology [@Skalski2005; @williams2002]. Over the last few decades, tremendous progress has been made towards this end. This progress is partly driven the development and application of new field data collection methods and approaches, such as citizen science data [@Danielsen2022], camera trap data [@Hamel2013] and the collection of environmental DNA data [@Beng2020]. In addition, developments of novel statistical methods alongside decreases in computational costs now allow researchers to estimate abundance and demographic rates in situations where it was not feasible before [@Zipkin2021]. Combined, these advances put us in a much better position for estimating quantities needed for population management [@williams2002] and indices relevant for large scale policy applications, e.g. Essential Biodiversity Variables [@Kissling2018].

Until recently, joint estimation of population dynamics and demography has relied mostly on data from marked individuals and associated open-population capture-mark-recapture models [@schaub2021]. While such methods can provide valuable information for both ecological research and management, collecting the necessary data is typically costly and logistically challenging to implement over large areas. Monitoring programmes focusing on abundance trends over larger areas, on the other hand, are typically based on data from unmarked animals. One often used approach for such surveys is to structure data collection around the distance sampling (DS) method. DS has been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015]. One reason for the method's popularity is that it does not require repeated visits to the same sites for estimating detection probability. This makes DS particularly useful for implementation in participatory monitoring programs, allowing stakeholders to take part in the data collection process.

The most common implementations of DS models have long used closed-population formulations, and thus do not allow formulating a process model that projects abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates @buckland2015. In recent years, DS approaches have been extended in many ways, including applications that estimate changes in abundance over time in open populations via a hidden state model representing population dynamics [@moore2011; @sollmann2015; @schaub2021]. This has greatly extended the potential of DS approaches for studying ecological dynamics across time and space. However, while these latter frameworks may allow to accurately quantify population changes, they typically provide little information on the drivers of these changes, i.e. the underlying vital rates. In fact, if the data does not contain information about the age (and/or sex-) structure of the surveyed population, there is no straightforward way to estimate demographic rates from such data. On the contrary, if age (and sex) of detected individuals can be determined, this information can be used to provide information on recruitment rates and survival probabilities. @Nilsen2018, for example, used a model based on harvest statistics and observations of population age structure to estimate population abundance and demographic rates without the need for any additional data from marked individuals.

@@

@@

The willow ptarmigan has a circumpolar distribution (Fuglei et al., 2020), and lives year-round in heterogeneous alpine and artic ecosystems. In Norway, there has been a long-term decline in the willow ptarmigan abundance across more than a century [@Hjeljord2022], but in the last few decades abundance trends has fluctuated both in time and space. In Scandinavia, willow ptarmigan is a valued game species (see e.g. @Andersen2014 ), and there have been several long-term research projects devoted to understanding how they respond to environmental variation and harvest management [@Israelsen2020; @Sandercock2011]. A key insight from across several study areas is the the annual recruitment rate (i.e. $R_t$ in our model, as outlined above) is highly variable, and is affected both by spring conditions [@Eriksen2023] and the abundance of small rodents, which constitute alternative prey for common predators (i.e. the Alternative Prey Hypothesis; see @Hagen1952; @Kausrud2008; @Bowler2020). Adult survival show less inter-annual fluctuations [@Israelsen2020], although spatial (and potentially temporal) variation due to e.g. harvest management is evident when comparing across studies [@Israelsen2020].

@@

Line transect survey data were collected in August each year, prior to the annual autumn harvest season, as part of the program "Hønsefuglportalen". Hønsefuglportalen is a national program for line transect surveys of tetraonid birds, and the effort is directed mainly towards willow ptarmigan habitats. In our case study, we used data from the western part of Lierne municipality. Line transects are surveyed by trained volunteers that use pointing dogs to locate the birds. When located, the geographical coordinate, perpendicular distance from the sampling line, the number of birds in the group, as well as the age (juvenile or adult) and sex of the birds are recorded. As the surveys are conducted in early August, juveniles can be distinguished from adults by their smaller body size. Males and females are manly distinguished by the sound (males often make a characteristic sound when being flushed. Nevertheless, mis-identification can occur, and in addition a proportion of the birds are registered as "unknown" age and/or sex. Since 2019, data has been collected through a mobile app tailormade for this project, which is available trhough App Store (for ios phones) and Google Play (for Android phones). Before 2019, field workers reported their data through a dedicated web portal. After data are collected and reported, they undergo several steps of quality control: first by local contacts and subsequently by personnel at the Norwegian Institute for Nature Research (NINA). The data are then standardized based on the Darwin-Core standard [@Wieczorek2012], and made publicly available as a sampling-event data set published through GBIF [@Nilsen2023a]. For additional description of the data collection procedures, see [@Bowler2020; @Kvasnes2018; @Nilsen2023a].

The individual longitudinal study based on radio collared willow ptarmigan was conducted in 2015-2021. Each winter (in February-March), willow ptarmigan were located at night using snowmobiles and large hand nets with prolonged handles, as described in [@Israelsen2020]. To prevent birds from flying off before the field personnel were close enough to capture them, a high-powered head lamp was used to dazzle the birds. After capture, birds were placed in an opaque bag to reduce stress. They were then fitted with a uniquely numbered leg ring (\~ 2.4g) and a Holohil RI-2BM or Holohil RI-2DM radio transmitter (\~ 14.1g) and subsequently released. The radio transmitters had an expected battery lifetime of 24 months (RI-2BM) or 30 months (RI-2DM), and included a mortality circuit that was activated if a bird had been immobile for 12 hours. We monitored the birds throughout most of the year by triangulation from the ground at least once a month for 10 months of the year (February -- November) by qualified field personnel. If a mortality signal was emitted from a transmitter, we attempted to recover it as soon as possible to determine cause of death. A number of birds dispersed out of the main study areas and was thus out of signal range for field personnel on the ground. To avoid loss of data, we conducted aerial triangulation using a helicopter or airplane three times a year (May, September and November) in the years 2016-2020.

@@

For the model fits to simulated and real data we ran 3 and 4 MCMC chains with NIMBLE's standard samples for 500k and 100k iterations, respectively. 300k and 40k and thereof were discarded as burn-in prior to thinning with factors 5 and 20 , leaving us with 40k and 3k posterior samples per chain (total of 120k and 12k samples per run), respectively. Posterior samples from the model fitted to real data are available at @Nilsen2024 (in folder PosteriorSamples_LierneCaseStudy).

@@

Posterior distributions for parameters estimated in three model fits to each of 10 simulated data sets are shown in @fig-sim-detection and @fig-sim-demo. Overall, the IDSM was able to correctly estimate both detection parameters and demographic rate parameters from all 10 simulated datasets without any systematic bias. The replicate runs for each dataset resulted in very similar posterior distributions, demonstrating that the models converged towards the same posterior distributions irrespective of starting values. Estimates of population sizes / densities were also adequate, and are presented in the supplementary materials (available at @Nilsen2024 in folder SimCheck_byDataSet).

@@

Given the available data, the IDSM was not able to estimate a clear effect of small rodent abundance on ptarmigan recruitment (slope-paramater for the z-standardized rodent occurrence data = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$Mean, 3)` ; 95% C.I. = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$lower, 3)` - `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$upper, 3)`, see supplementary figure "Rep_betaR.R.png" in @Nilsen2024 ).

@@

We developed an integrated populaiton model that jointly analyses line transect distance sampling survey data and data from marked individuals to estimate population abundance, survival probabilities, and recruitment rates over time. We first used simulated data to examine the model's ability to recover the underlying parameters when they were known. We then fitted the model to data from an ongoing field study on willow ptarmigan in Norway to showcase its applicability to real wildlife monitoring data.

Open population formulations of the distance sampling model have previously been presented [@sollmann2015; @Bowler2020; @moore2011], and in these studies also applied to various ecological systems. The model presented here extends those previous applications by formulating the underlying population process model as a stage-structured matrix model [@caswell2000] in which the matrix elements are represented by annual survival probabilities and recruitment rates. While this has been the common approach for a range of other statistical modelling frameworks, including the growing suite of models falling into the category of integrated population models [@schaub2021], the integration of mechanistic population models into distance sampling frameworks is rather new. The resulting modelling framework allow us to make maximum use of distance sampling data in combination with auxiliary information both from the distance sampling survey itself (i.e. information on age, sex, etc. of observed animals) and from other types of monitoring, and enables estimation not only of changes in population density but also of underlying vital rates over time.

@@

Real data, however, are likely to be subject to certain biases, and such biases might result in biased parameter estimates unless accounted for. One type of bias that is likely to be common for age-structured distance sampling data arises from failure to (correctly) classify the age class of observed individuals. In our case study on willow ptarmigan in Norway, such misclassification is likely to happen at an unknown rate, even if the size difference between adult and juvenile birds are quite substantial during the survey. Moreover, The probability for misclassificiation might be related to both the timing of the survey (e.g. mid August rather than early August), it might vary between observers, and even by survey conditions. Observations with incorrectly classified age have the potential to introduce bias in the IDSMs relative estimates of survival and recruitment. This is due to the way it useses the distance sampling data to estimate survival and recruitment rates. In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, and this creates a dependence between the demographic parameters. If the age ratio in the data are biased or contain frequent misclassifications, this is likely to affect the relative contribution of survival and recruitment to the growth rate. To get an idea of the potential effect of this on parameter estimation, we checked the sensitivity of the output of the model fit to real data with regard to the treatment of birds classified as "unknown sex and age" by the field personnel (see Methods). In the model version presented in the results section, we made the assumption that these birds were in fact juveniles. Comparing estimates to an alternative scenario in which we discarded all birds classified as "unknown sex and age" (see supplementary figures in @Nilsen2024 ) we found that -- as expected -- estimated population density was virtually unaffected by the treatment of "unknown sex and age" observations, while the annual demographic parameters shifted proportional to the amount of "unknown sex and age" observations in the given year (towards higher recruitment and lower survival). Thus, biases in the reported age ratios may affect estimated of demographic rates, but not so much population density. Since the proportion of "unknown sex and age" observations in our ptarmigan case study was rather low, potential biases in estimates resulting from age misclassification are expected to be small. Nonetheless, future developments of the IDSM modelling framework should focus on ways of accounting explicitly for misclassification of age class in the field.

The density estimates that we derived from the case study on Willow ptarmigan in Norway is comparable to previous estimates from across Norway (see e.g. @Sandercock2011; @Kvasnes2014b). Throughout the study period from 2007 - 2021, the density increased markedly, but the reason for this increase is not known. Compared to previous studies on ptarmigan (see e.g. [@Israelsen2020; @Sandercock2011]), we could have expected somewhat higher estimates of survival probability. One potential reason for the overall lower estimates obtained here is that our IDSM analysis assumed constant survival over the period 2007-2021 while survival, in reality, may have changed over time. If survival in more recent years, when telemetry data was collected (and the study of e.g. @Israelsen2020 was carried out) was higher than in earlier years -- something that seems likely given population increase over recent years -- an average over the entire time period is expected to be lower. On a different note, we can also not exclude the possibility of a small degree of bias in survival estimates due to misclassification of age in the data (see above), especially seeing as the IDSM's recruitment rate estimates also appear somewhat high compared to other studies on willow ptarmigan [@Eriksen2023; @Kvasnes2014a; @Steen1988]. Note however that the last period of the study period in Lierne, there was substantial increase in the density (ref Figure 2), which might indicate that the recruitment and/or survival rates were high in this period.

In addition to estimating demographic rates from line transect data, the IDSM also allows including relevant environmental effects on the demographic rates themselves, and not just on population growth rate as a whole ($\lambda$). In the ptarmigan case study we thus attempted to investigate the effect of small rodent abundance (approximated as the proportion of transect lines on which rodents were reported each year) on recruitment rate. Unfortunately, we were not able to detect a clear effect of rodent abundance due to large uncertainty associated with the estimate (see supplementary figures in @Nilsen2024). This may seem somewhat surprising given that such a pattern has been reported repeatedly in the literature (see e.g. @Bowler2020). We speculate that there are at least three potential (non-mutual) explanations to this result. The first is that our covariate data may not have been well suited for estimating effects on recruitment. The data on rodent abundance was heavily zero-inflated, and the annual variation in the index was rather small otherwise, making for a covariate with relatively little information content. While this may be partially a consequence of how these data are collected, it is also well known that the amplitude and regularity of the rodent cycles has been fading in recent decades [@Kausrud2008; @Cornulier2013], and our study area might be no exception. Lack of peak rodent years in the time series to which we fitted the model may thus also have contributed to making effect estimation challenging. Second, it is possible that rodent effects were obscured by other, potentially stronger, covariate effects. Previous research has shown that ptarmigan recruitment is also sensitive to the weather in the late winter and spring (before and during the breeding season); as we did not fit any weather covariates to the model so there is a possibility that strong unaccounted for effects of spring conditions in certain years may have masked any remaining effects of small rodent abundance. Finally, the data set used in this analysis is relatively short (15 years), leaving us with somewhat limited statistical power to detect effects of temporal covariates . Taken together we therefore do not consider this study as a particularly strong test of the underlying effect of small rodent fluctuations on ptarmigan recruitment rates. It is worth noting that future applications could increase statistical power by including either more years of data or capitalizing on space-for-time substitution as the Norwegian ptarmigan monitoring programme spans many more locations beyond Lierne. @Bowler2020, for example, used data from the same sampling program but from more areas using a simpler open population DS model, and detected a very clear signal from small rodent abundance on ptarmigan population growth rate. An extension of our IDSM to include data from multiple areas therefore constitutes a promising approach for investigating to which extent similar results emerge by linking environmental covariates to the actual demographic rates and not only just to the resulting population growth rates.

@@

We are grateful to field workers that collected the line transect data through the Hønsefuglportalen program. Fjellstyrene i Lierne and many colleagues from NINA and Nord University contributed both to the line transect program and the field work related to the marked willow ptarmigan. We also thank James A. Martin for constructive discussions during the later stages of this work.

### Funding

Both data collection and analytical work were financed by the Norwegian Environment Agency funded the projects (grant nrs 17010522, 19047014 and 22047004).

### Conflict of interest disclosure

The authors declare that they comply with the PCI rule of having no financial conflicts of interest in relation to the content of the article

@@

The raw data from the line transect surveys is deposited on GBIF and can be accessed freely via the Living Norway Data Portal (https://data.livingnorway.no). The work here is based on version 1.7 of the dataset "Tetraonid line transect surveys from Norway: Data from Fjellstyrene" [@Nilsen2023a].

The auxiliary radio-telemetry data and rodent occupancy data, and all code used for wrangling, analysing, and visualizing data and results can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>. The results presented in this paper were created using version 1.3 of the code [@IDSMCode_2024].

Supplementary figures and posterior samples from the model run on the real data are available as a time-stamped open archived on Open Science Framework [@Nilsen2024].

## References
